*Supplement* Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary

### 2024-05-07 manu-draft.qmd
add supp gam figs and tabs

The long-term recovery of seagrass habitats in Tampa Bay since the 1980s is a nationally recognized success story that demonstrates application of a successful management paradigm through the EPA-administered National Estuary Program [@greening2006; @greening2014; @sherwood2017]. From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of a benchmark 1950s coverage. Though Tampa Bay was far from pristine in the 1950s, aerial imagery was sufficient to estimate a relatively unimpacted condition for seagrass coverage. Since that time, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a). Light environments have improved (Figure [-@fig-seagrasschg]c) through reduction of external nitrogen loadings by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed \[@greening2014; @beck2019\].

@@

```{r}
# secchi on bottom is na issue
# also averaging all models by year in the bay segment for comparable sample size, fim/pdem limited to july - nov to minimize seasonal effect and when transect surveys completed for comparability
# < 17% for epc, much higher for fim/pdem so la not included, but can show la relatively stable in a plot
# epc is bonk becuase of disconnect between sites
# fim is good, makes sense, but otb is weird
# pdem temp makes sense for otb
```

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the annual summaries. As such, the annual stressor metrics were compared to frequency occurrence (all species) each year by bay segment. GAMs were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, or both. Additional predictors included year and light attenuation as estimated from Secchi depth [@tbep0696]. For 1998 to 2022, `r onb$epc` of the Secchi observations were recorded on the bottom. A single smooth term was used for each predictor using a thing plate regression spline, including a tensor product smoother term that evaluated the potential interacting effects of each predictor with year. Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric. All models excluded Lower Tampa Bay because of minimal seagrass change over time. Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by *H. wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b).

Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change. These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location). GAMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables. Interactions with year for each predictor variable and separate smoothed terms for each bay segment were also included as above. GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data. Neither the FIM nor PDEM models used light attenuation as a predictor variable given that a large percentage of Secchi observations were measured on the bottom (`r onb$fim` and `r onb$pinco`, respectively), providing further support that light environments have not been limiting for seagrasses. Model input data were further subset to include only months from July to October to describe seagrasses during the growing season and to reduce potential seasonal effects. Lastly, all data were averaged annually for the monthly subsets for comparability of sample size (i.e., power) with the EPC models. For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs.

Add equations.....

@@

```{r}
# epcmodsum1 <- modtxt_fun(sgmods$epcmod1)
# epcmodsum2 <- modtxt_fun(sgmods$epcmod2)
# fimmodsum <- modtxt_fun(sgmods$fimmod)
```

Linear models to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass loss after 2016 was influenced by climate stressors. The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (). An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab). For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30$^\circ$ increased, whereas no relationship was observed after 2016. However, the interaction between time period and temperature was not significant, suggesting no difference between time periods. The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab). The interaction between the temperature and salinity metric was included in the most parsimonious model, although it was not significant. The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab). Note that [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down.

>>>>

The long-term recovery of seagrass habitats in Tampa Bay since the 1980s is a nationally recognized success story that demonstrates application of a successful management paradigm through the EPA-administered National Estuary Program [@greening2006; @greening2014; @sherwood2017]. From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of a benchmark 1950s coverage. Though Tampa Bay was far from pristine in the 1950s, aerial imagery was sufficient to estimate a relatively unimpacted condition for seagrass coverage. Since that time, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a). Light environments have improved (Figure [-@fig-seagrasschg]c) through reduction of external nitrogen loadings by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].

@@

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the annual summaries. As such, the annual stressor metrics were compared to frequency occurrence (all species) each year by bay segment. GAMs were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, or both. Additional predictors included year and light attenuation as estimated from Secchi depth [@tbep0696]. For 1998 to 2022, `r onb$epc` of the Secchi observations were recorded on the bottom. A single smooth term was used for each predictor using a thin plate regression spline, including a tensor product interaction term with a cubic regression spline that evaluated the potential interacting effects of each predictor with year. Two models were evaluated, one with the temperature and salinity metrics together (@eq-epcmod1) and another with the both metric (@eq-epcmod2). All models excluded Lower Tampa Bay because of minimal seagrass change over time. Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by *H. wrightii* in the upper bay segments (Figure [-@fig-seagrasschg]b). The GAM equations in mgcv R package notation [@wood17] are below, where $s$ is an smoother function for a single predictor and $ti$ is the tensor product interaction between predictors. All GAMs were fit using restricted maximum likelihood evaluation [@wood11].

$$
\begin{aligned}
Seagrass \sim s(year, by = bay \ segment) + s(light \ att, by = bay \ segment) + 
\\
s(temperature, by = bay \ segment) + s(salinity, by = bay \ segment) + 
\\
ti(light \ att, \ year, by = bay \ segment) + ti(temperature, year, by = bay \ segment) +
\\ 
ti(salinity, year, by = bay \ segment)
\end{aligned}
$$ {#eq-epcmod1}

$$
\begin{aligned}
Seagrass \sim s(year, by = bay \ segment) + s(light \ att, by = bay \ segment) + 
\\
s(both, by = bay \ segment) +
\\
ti(light \ att, \ year, by, bay \ segment) + ti(both, year, by = bay \ segment)
\end{aligned}
$$ {#eq-epcmod2}
Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change. These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location). GAMs were used for the FIM data to evaluate seagrass percent cover using year, temperature, and salinity as independent variables. Separate smoothed terms for each predictor and iteractions with year for each predictor for each bay segment were included as above. GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data. Neither the FIM nor PDEM models used light attenuation as a predictor variable given that a large percentage of Secchi observations were measured on the bottom (`r onb$fim` and `r onb$pinco`, respectively), providing further support that light environments have not been limiting for seagrasses. Model input data were further subset to include only months from July to October to describe seagrasses during the growing season and to reduce potential seasonal effects. Lastly, all data were averaged annually for the monthly subsets for comparability of sample size (i.e., power) with the EPC models. For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs. The GAM equations for the FIM (@eq-fimmod) and PDEM (@eq-pincomod) datasets were as follows.

$$
\begin{aligned}
Seagrass \sim s(year, by = bay \ segment) + s(temperature, by = bay \ segment) + 
\\
s(salinity, by = bay \ segment) + 
\\
ti(temperature, year, by = bay \ segment) + ti(salinity, year, by = bay \ segment)
\end{aligned}
$$ {#eq-fimmod}

$$
\begin{aligned}
Seagrass \sim s(year) + s(temperature) + s(salinity) + 
\\
ti(temperature, year) + ti(salinity, year)
\end{aligned}
$$ {#eq-pincomod}

@@

```{r}
# epcmodsum1 <- modtxt_fun(sgmods$epcmod1)
# epcmodsum2 <- modtxt_fun(sgmods$epcmod2)
# fimmodsum <- modtxt_fun(sgmods$fimmod)
# secchi on bottom is na issue
# also averaging all models by year in the bay segment for comparable sample size, fim/pdem limited to july - nov to minimize seasonal effect and when transect surveys completed for comparability
# < 17% for epc, much higher for fim/pdem so la not included, but can show la relatively stable in a plot
# epc is bonk becuase of disconnect between sites
# fim is good, makes sense, but otb is weird
# pdem temp makes sense for otb
```

GAMs to assess the potential effects of temperature and salinity on seagrass change provided modest evidence that seagrass change was influenced by climate stressors. The most parsimonious model for the EPC data included the temperature and salinity metrics as potentially important for explaining inter-annual variation in seagrass frequency occurrence (). An interaction of the metrics with time period was also observed, with notable differences by predictor (Figure [-@fig-sgmod]a, @tbl-modtab). For the period prior to 2016, frequency occurrence of seagrass increased as the number of days where temperature was above 30$^\circ$ increased, whereas no relationship was observed after 2016. However, the interaction between time period and temperature was not significant, suggesting no difference between time periods. The opposite trend was observed for the number of days when salinity was below 25 ppt, where no relationship was observed prior to 2016 and a decreasing relationship was observed after 2016 (interaction term between salinity and time period was marginally significant, p \< 0.1, Figure [-@fig-sgmod]a, @tbl-modtab). The interaction between the temperature and salinity metric was included in the most parsimonious model, although it was not significant. The second model for the EPC data that evaluated the number of days when both temperature was above and salinity was below the threshold was also significant (), although the interaction term with time period was not included in the most parsimonious model, suggesting no effect of this metric in describing seagrass change after 2016 (@tbl-modtab). Note that [@fig-sgmod] shows model predictions for OTB only, although the shape of the relationships do not change for other bay segments other than as additive shifts of the curves up or down.

### 2024-05-07 supplement.qmd
add supp gam figs and tabs

load(file = here('tabs/supptempfittab.RData'))
load(file = here('tabs/suppsalifittab.RData'))
load(file = here('tabs/supp1mixtab.RData'))
load(file = here('tabs/supp2mixtab.RData'))
load(file = here('tabs/supp1daytab.RData'))
load(file = here('tabs/supp2daytab.RData'))
```

@@

<!-- ![Relationship between water temperature and discharge in the Hillsborough River (USGS gage 02303000) over time. The relationship is shown for February (dry month) and August (wet month) grouped by historic (1975-1999) and contemporary (2000-2022) time periods. Regression lines with 95% confidence intervals for each temporal grouping are shown.](figs/suppflowtemp.png){#fig-suppflowtemp} -->

<!-- {{< pagebreak >}} -->

@@

@@

>>>>

load(file = here('tabs/supptempfittab.RData'))
load(file = here('tabs/suppsalifittab.RData'))
load(file = here('tabs/supp1mixtab.RData'))
load(file = here('tabs/supp2mixtab.RData'))
load(file = here('tabs/supp1daytab.RData'))
load(file = here('tabs/supp2daytab.RData'))
load(file = here('tabs/suppepcmod1tab.RData'))
load(file = here('tabs/suppepcmod2tab.RData'))
load(file = here('tabs/suppfimmodtab.RData'))
load(file = here('tabs/supppincomodtab.RData'))
```

@@

@@

![Partial effects of smoothers by bay segment from a Generalized Additive model used to describe seagrass change relative to year, light attenuation, the number of days each year when temperature was above 30 $^\circ$C, and the number of days each year when salinity was below 25 ppt. The EPC data were used to develop the model. Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors. The model also included smoothed interaction terms with year (not shown). See Table S7 for additional model fit statistics. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay. n = 75, R.sq. 0.93, Deviance explained 97\%](figs/suppepcmod1.png){#fig-suppepcmod1}
{{< pagebreak >}}

![Partial effects of smoothers by bay segment from a Generalized Additive model used to describe seagrass change relative to year, light attenuation, and the number of days each year when both temperature was above 30 $^\circ$C and salinity was below 25 ppt. The EPC data were used to develop the model. Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors. The model also included smoothed interaction terms with year (not shown). See Table S8 for additional model fit statistics. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay. n = 75, R.sq. 0.93, Deviance explained 97%](figs/suppepcmod2.png){#fig-suppepcmod2}
{{< pagebreak >}}

![Partial effects of smoothers by bay segment from a Generalized Additive model used to describe seagrass change relative to year, temperature ($^\circ$C), and salinity (ppt). The FIM data were used to develop the model. Partial effects describe the modeled association between each predictor and mean seagrass coverage (0-100) after accounting for the effects of the other predictors. The model also included smoothed interaction terms with year (not shown). See Table S9 for additional model fit statistics. OTB: Old Tampa Bay, HB: Hillsborough Bay, MTB: Middle Tampa Bay. n = 81, R.sq. 0.81, Deviance explained 89%](figs/suppfimmod.png){#fig-suppfimmod}
{{< pagebreak >}}

![Partial effects of smoothers from a Generalized Additive model used to describe seagrass change relative to year, temperature ($^\circ$C), and salinity (ppt). The PDEM data in Old Tampa Bay were used to develop the model. Partial effects describe the modeled association between each predictor and seagrass frequency occurrence after accounting for the effects of the other predictors. The model also included smoothed interaction terms with year (not shown). See Table S10 for additional model fit statistics. n = 19, R.sq. 0.69, Deviance explained 81%](figs/supppincomod.png){#fig-supppincomod}

@@

```{r}
#| label: tbl-suppepcmod1tab
suppepcmod1tab
```

```{r}
#| label: tbl-suppepcmod2tab
suppepcmod2tab
```

```{r}
#| label: tbl-suppfimmodtab
suppfimmodtab
```

```{r}
#| label: tbl-supppincomodtab
supppincomodtab
```
